---
title: "500 Lab 0 Template: CHANGE THIS TITLE"
author: "YOUR NAME GOES HERE"
date: last-modified
format: 
  html:
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    date-format: iso
    theme: default  ## change the theme if you prefer
---

# R Setup {.unnumbered}

```{r}
#| message: false
#| warning: false

knitr::opts_chunk$set(comment = NA) # do not remove this

library(janitor) # load other packages as desired
library(tidyverse) # load tidyverse last

theme_set(theme_bw()) # set theme for ggplots
```

## Loading the dig1 Data {.unnumbered}

```{r}
dig1 <- read_csv("https://raw.githubusercontent.com/THOMASELOVE/500-data/master/data/dig1.csv",
                 show_col_types = FALSE) |>
  mutate(subjectid = as.character(subjectid)) |>
  clean_names()

dig1
```

Note the `dig1` tibble should have 6800 rows and 72 columns.

# Get access to the DIG training data

> Visit <https://biolincc.nhlbi.nih.gov/teaching/> and request copies of the DIG "teaching" data set.

Verify that you've done this, in a complete sentence or two.

# Mock proposal for a DIG study

> Review the documentation available to you on the BIOLINCC page, and complete a **mock proposal** not to exceed 300 words for an observational study based on the available data from the DIG randomized trial. A list of questions you should address follows below. 

## What comparison do you want to make?

Response goes here.

## Why is this of interest?

Response goes here.

## What are the key measures available in the data?

Response goes here.

# Modeling the DIG data

## Preparing the Sample for Modeling

```{r}
set.seed(5002023)
dig_hw1 <- dig1 |>
    mutate(nyha = factor(functcls),
           female = sex - 1,
           race_f = fct_recode(factor(race), 
                               "White" = "1",
                               "Non-White" = "2"),
           tx_f = fct_recode(factor(trtmt), 
                                    "Placebo" = "0",
                                    "Treatment" = "1"),
           death_f = fct_recode(factor(death),
                                "Died" = "1",
                                "Survived" = "0")) |>
    select(subjectid, death_f, tx_f, age, race_f, female,
           ejf_per, bmi, nyha, angina) |>
    slice_sample(n = 5000)
```

## Initial Data Summaries

## Partitioning the Data into Training and Test Samples

## Fitting a Logistic Regression Model to the training sample

## Applying the model to the test sample, and summarizing results

## Conclusions





